---
title: "ME/CFS and the Oral Microbiome"
author: "Sean Bramley"
date: "2/15/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(knitr)
library(plyr)
library(tidyverse)
library(gdata)
library(gridExtra)
library(ggplot2)
library(lme4) 
library(lmerTest)
library(mgcv) 
library(metamicrobiomeR)
library(Hmisc)
library(Gmisc)
library(sqldf)
library(caret)
#install.packages("arsenal")
library(arsenal)

#install.packages("gamlss")

library(gamlss)
```

##Read in Data
```{r read in}
metadata <- read_csv("C:/Users/seanl/Desktop/Spring 2021/Thesis/Analysis/Data/RSA_metadataOA_2021_1_20.186_cheng.csv") %>%
  janitor::clean_names() %>%
  mutate(sample_type = factor(sample_type, c("Control", "Case")),
         sex = factor(sex, c("F", "M")),
         bmi_cat = factor(bmi_cat, c("Underweight", "Normal", "Overweight", "Obese")),
         ibs_dx = factor(ibs_dx, c("Y", "N")),
         allergy_dx = factor(allergy_dx, c("Y", "N")),
         anti_depressant_rx = factor(anti_depressant_rx, c("Y", "N")),
         pain_narc_rx = factor(pain_narc_rx, c("Y", "N")),
         antimicro_rx = factor(antimicro_rx, c("Y", "N")),
         antibiotic_use = factor(antibiotic_use, c("Y", "N")),
         probiotic_supp_use = factor(probiotic_supp_use, c("Y", "N")),
         prebiotic_supp_use = factor(prebiotic_supp_use, c("Y", "N")),
         site = factor(site, c("california", "florida", "nevada", "utah", "new_york")),
         category = factor(category, c("MECFS_IBS", "MECFS_woIBS", "Control", "Control_IBS")),
         race_hispanic = factor(race_hispanic, c("White_Non_Hispanic", "Hispanic", "Non_White_Non_Hispanic"))
         )

alpha_shannon <- read_csv("C:/Users/seanl/Desktop/Spring 2021/Thesis/Analysis/Data/shannon.csv") %>%
  janitor::clean_names() %>%
  dplyr::select(saliva_id, shannon)
alpha_chao1 <- read_csv("C:/Users/seanl/Desktop/Spring 2021/Thesis/Analysis/Data/chao1.csv") %>%
  janitor::clean_names() %>%
  dplyr::select(saliva_id, chao1)
alpha_evenness <- read_csv("C:/Users/seanl/Desktop/Spring 2021/Thesis/Analysis/Data/evenness.csv") %>%
  janitor::clean_names() %>%
  dplyr::select(saliva_id, evenness)
alpha_otus <- read_csv("C:/Users/seanl/Desktop/Spring 2021/Thesis/Analysis/Data/observed_otu.csv") %>%
  janitor::clean_names() %>%
  dplyr::select(saliva_id, observed_otus)
```

##Joining Alpha Diversity Data sets to the Metadata file
```{r sql}
alpha_merge <- sqldf("SELECT a.shannon, b.chao1, c.evenness, d.observed_otus, m.*
                     FROM metadata AS m
                     INNER JOIN alpha_shannon AS a
                     ON m.saliva_id = a.saliva_id
                     INNER JOIN alpha_chao1 AS b
                     ON m.saliva_id = b.saliva_id
                     INNER JOIN alpha_evenness AS c
                     ON m.saliva_id = c.saliva_id
                     INNER JOIN alpha_otus AS d
                     ON m.saliva_id = d.saliva_id;", method="raw")

alpha_merge$race_hispanic <- factor(alpha_merge$race_hispanic, c("White_Non_Hispanic", "Hispanic", "Non_White_Non_Hispanic"))
alpha_merge$sex <- factor(alpha_merge$sex, c("M", "F"))


write.csv(alpha_merge, "C:/Users/seanl/Desktop/Spring 2021/Thesis/Analysis/Data/alpha.merge.csv")
#Checking distribution of alpha diversity
hist(alpha_merge$shannon)
hist(alpha_merge$chao1)
hist(alpha_merge$evenness)
hist(alpha_merge$observed_otus)
#all approximately normal 
```


#Comparing model fits 
```{r alpha lm}
alpha_models <- alpha_merge %>%
    mutate(bmi = scale(bmi),
         age_at_a1 = scale(age_at_a1)) %>%
  mutate(bmi = as.double(bmi),
         age_at_a1 = as.double(age_at_a1))

lm_shannon <- lm(shannon ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx, data = alpha_models)
lm_chao1 <- lm(chao1 ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx, data = alpha_models)
lm_evenness <- lm(evenness ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx, data = alpha_models)

lm_ibs <- lm(shannon ~ ibs_dx, data = alpha_models)
lm_mecfs <- lm(shannon ~ mecfs_status, data = alpha_models)
lm_mecfsibs <- lm(shannon ~ mecfs_status + ibs_dx, data = alpha_models)
summary(lm_mecfsibs)
summary(lm_mecfs)
summary(lm_ibs)
summary(lm_shannon)
summary(alpha_models$shannon)
table(alpha_models$ibs_dx)
table(alpha_models$mecfs_status)
BIC(lm_shannon) #500.3964
BIC(lm_chao1) #2208.77
BIC(lm_evenness) #-262.9123

hist(alpha_models$shannon)
```

```{r negative binomial for count}
otu_nb <- glm.nb(observed_otus ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + ibs_dx, maxit = 100, data = alpha_models)
summary(otu_nb)
BIC(otu_nb) #2201.204
```

Compared linear models with GAMLSS and GLM with gamma dist. Linear models fit best for shannon, chao1, and evenness. OTU's modeled with glm poisson-link and negative binomial. NB fit best for OTU. 

#Checking for confounding in linear models
```{r confounding}
painrx_shannon <- lm(shannon ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + pain_narc_rx, data = alpha_models)
painrx_chao1 <- lm(chao1 ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + pain_narc_rx, data = alpha_models)
painrx_evenness <- lm(evenness ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + pain_narc_rx, data = alpha_models)

summary(lm_shannon) #MECFS: -0.205
summary(lm_chao1) #MECFS: -4.630
summary(lm_evenness) #MECFS: -0.018

summary(painrx_shannon) #MECFS: -0.118
summary(painrx_chao1) #MECFS: -3.931
summary(painrx_evenness) #MECFS: -0.008

(-0.205--0.118)/-0.118 #73.72% change
(-4.630--3.931)/-3.931 #17.78% change
(-0.018--0.008)/-0.008 #125% change 
```

```{r antidepressant}
depressrx_shannon <- lm(shannon ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + anti_depressant_rx, data = alpha_models)
depressrx_chao1 <- lm(chao1 ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + anti_depressant_rx, data = alpha_models)
depressrx_evenness <- lm(evenness ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + anti_depressant_rx, data = alpha_models)

summary(lm_shannon) #MECFS: -0.205
summary(lm_chao1) #MECFS: -4.630
summary(lm_evenness) #MECFS: -0.018

summary(depressrx_shannon) #MECFS: -0.228
summary(depressrx_chao1) #MECFS: -3.171
summary(depressrx_evenness) #MECFS: -0.020

(-0.205--0.228)/-0.228 #10.09% change
(-4.630--3.171)/-3.171 #46.07% change
(-0.018--0.020)/-0.020 #10% change 
```

Anti-depressant/pain narcotics appear to be confounders so I ran adjusted models
```{r adjusted models}
adjusted_shannon <- lm(shannon ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + anti_depressant_rx + pain_narc_rx, data = alpha_models)
adjusted_chao1 <- lm(chao1 ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + anti_depressant_rx + pain_narc_rx, data = alpha_models)
adjusted_evenness <- lm(evenness ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + ibs_dx + anti_depressant_rx  + pain_narc_rx, data = alpha_models)
adjusted_otu <- glm.nb(observed_otus ~ mecfs_status + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + ibs_dx + anti_depressant_rx + pain_narc_rx, maxit = 100, data = alpha_models)

summary(lm_shannon) #MECFS: -0.205
summary(lm_chao1) #MECFS: -4.630
summary(lm_evenness) #MECFS: -0.018
summary(otu_nb) #MECFS: -0.019

summary(adjusted_shannon) #MECFS: -0.143
summary(adjusted_chao1) #MECFS: -2.422
summary(adjusted_evenness) #MECFS: -0.0103
summary(adjusted_otu) #MECFS: -0.014

(-0.205--0.143)/-0.143 #43.36% change
(-4.630--2.422)/-2.422 #91.16% change
(-0.018--0.0103)/-0.0103 #75% change 
(-0.019--0.014)/-0.014 #35.71% change
```

Alpha diversity by state
```{r state linear models}
ny_shannon <- alpha_models %>%
  filter(site == "new_york")

lm_ny_shannon <- lm(shannon ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = ny_shannon)
lm_ny_chao1 <- lm(chao1 ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = ny_shannon)
lm_ny_evenness <- lm(evenness ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = ny_shannon)
lm_ny_otu <- glm.nb(observed_otus ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, maxit = 100, data = ny_shannon)

summary(lm_ny_shannon)
summary(lm_ny_chao1)
summary(lm_ny_evenness)
summary(lm_ny_otu)

nv_alpha <- alpha_models %>%
  filter(site == "nevada")

lm_nv_shannon <- lm(shannon ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = nv_alpha)
lm_nv_chao1 <- lm(chao1 ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = nv_alpha)
lm_nv_evenness <- lm(evenness ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = nv_alpha)
lm_nv_otu <- glm.nb(observed_otus ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, maxit = 100, data = nv_alpha)

summary(lm_nv_shannon)
summary(lm_nv_chao1)
summary(lm_nv_evenness)
summary(lm_nv_otu)

ca_alpha <- alpha_models %>%
  filter(site == "california")
table(ca_alpha$pain_narc_rx)

lm_ca_shannon <- lm(shannon ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx, data = ca_alpha)
lm_ca_chao1 <- lm(chao1 ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx, data = ca_alpha)
lm_ca_evenness <- lm(evenness ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx, data = ca_alpha)
lm_ca_otu <- glm.nb(observed_otus ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx, maxit = 100, data = ca_alpha)

summary(lm_ca_shannon)
summary(lm_ca_chao1)
summary(lm_ca_evenness)
summary(lm_ca_otu)

ut_alpha <- alpha_models %>%
  filter(site == "utah")
table(ut_alpha$prebiotic_supp_use)

lm_ut_shannon <- lm(shannon ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = ut_alpha)
lm_ut_chao1 <- lm(chao1 ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = ut_alpha)
lm_ut_evenness <- lm(evenness ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = ut_alpha)
lm_ut_otu <- glm.nb(observed_otus ~ mecfs_status + ibs_dx + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use + probiotic_supp_use + anti_depressant_rx + pain_narc_rx, maxit = 100, data = ut_alpha)

summary(lm_ut_shannon)
summary(lm_ut_chao1)
summary(lm_ut_evenness)
summary(lm_ut_otu)

```

Alpha diversity in PAM clusters
```{r cluster alpha diversity}

new_metadata <- readr::read_tsv("C:/Users/seanl/Desktop/Spring 2021/Thesis/Analysis/Data/RSA_metadataOA_2021_3_23.186.txt", col_names=TRUE) %>%
  janitor::clean_names() %>%
  mutate(sample_type = factor(sample_type, c("Control", "Case")),
         sex = factor(sex, c("F", "M")),
         bmi_cat = factor(bmi_cat, c("Underweight", "Normal", "Overweight", "Obese")),
         ibs_dx = factor(ibs_dx, c("Y", "N")),
         allergy_dx = factor(allergy_dx, c("Y", "N")),
         anti_depressant_rx = factor(anti_depressant_rx, c("Y", "N")),
         pain_narc_rx = factor(pain_narc_rx, c("Y", "N")),
         antimicro_rx = factor(antimicro_rx, c("Y", "N")),
         antibiotic_use = factor(antibiotic_use, c("Y", "N")),
         probiotic_supp_use = factor(probiotic_supp_use, c("Y", "N")),
         prebiotic_supp_use = factor(prebiotic_supp_use, c("Y", "N")),
         race_hispanic = factor(race_hispanic, c("White_Non_Hispanic", "Hispanic", "Non_White_Non_Hispanic")),
         site = factor(site, c("california", "florida", "nevada", "utah", "new_york")),
         category = factor(category, c("MECFS_IBS", "MECFS_woIBS", "Control", "Control_IBS")),
         pam2 = factor(pam2, c("1", "2"))
  )

alpha_cluster_merge <- sqldf("SELECT a.shannon, b.chao1, c.evenness, d.observed_otus, m.*
                     FROM new_metadata AS m
                     INNER JOIN alpha_shannon AS a
                     ON m.saliva_id = a.saliva_id
                     INNER JOIN alpha_chao1 AS b
                     ON m.saliva_id = b.saliva_id
                     INNER JOIN alpha_evenness AS c
                     ON m.saliva_id = c.saliva_id
                     INNER JOIN alpha_otus AS d
                     ON m.saliva_id = d.saliva_id;", method="raw")

alpha_cluster_merge$race_hispanic <- factor(alpha_cluster_merge$race_hispanic, c("White_Non_Hispanic", "Hispanic", "Non_White_Non_Hispanic"))
alpha_cluster_merge$sex <- factor(alpha_cluster_merge$sex, c("M", "F"))

c1_alpha <- alpha_cluster_merge %>%
  filter(pam2 == "1")

lm_c1_shannon <- lm(shannon ~ mecfs_status + ibs_dx + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = c1_alpha)
lm_c1_chao1 <- lm(chao1 ~ mecfs_status + ibs_dx + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = c1_alpha)
lm_c1_evenness <- lm(evenness ~ mecfs_status + ibs_dx + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = c1_alpha)
lm_c1_otu <- glm.nb(observed_otus ~ mecfs_status + ibs_dx + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, maxit = 100, data = c1_alpha)

summary(lm_c1_shannon)
summary(lm_c1_chao1)
summary(lm_c1_evenness)
summary(lm_c1_otu)

c2_alpha <- alpha_cluster_merge %>%
  filter(pam2 == "2")

lm_c2_shannon <- lm(shannon ~ mecfs_status + ibs_dx + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = c2_alpha)
lm_c2_chao1 <- lm(chao1 ~ mecfs_status + ibs_dx + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = c2_alpha)
lm_c2_evenness <- lm(evenness ~ mecfs_status + ibs_dx + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use +
                   probiotic_supp_use + prebiotic_supp_use + anti_depressant_rx + pain_narc_rx, data = c2_alpha)
lm_c2_otu <- glm.nb(observed_otus ~ mecfs_status + ibs_dx + site + sex + bmi + race_hispanic + age_at_a1 + antibiotic_use + probiotic_supp_use + prebiotic_supp_use  + anti_depressant_rx + pain_narc_rx, maxit = 100, data = c2_alpha)

summary(lm_c2_shannon)
summary(lm_c2_chao1)
summary(lm_c2_evenness)
summary(lm_c2_otu)

```

Output alpha diversity models as tables
```{r final alpha models}
library(sjPlot)

tab_model(adjusted_shannon, adjusted_chao1, adjusted_evenness, adjusted_otu)
tab_model(lm_shannon, lm_chao1, lm_evenness, otu_nb)

tab_model(lm_ny_shannon, lm_ny_chao1, lm_ny_evenness, lm_ny_otu)
tab_model(lm_nv_shannon, lm_nv_chao1, lm_nv_evenness, lm_nv_otu)
tab_model(lm_ca_shannon, lm_ca_chao1, lm_ca_evenness, lm_ca_otu)
tab_model(lm_ut_shannon, lm_ut_chao1, lm_ut_evenness, lm_ut_otu)

tab_model(lm_c1_shannon, lm_c1_chao1, lm_c1_evenness, lm_c1_otu)
tab_model(lm_c2_shannon, lm_c2_chao1, lm_c2_evenness, lm_c2_otu)
```

Logistic regression models for FDAARGOS 522
```{r fdaargos522 logistic}
daargos <- readxl::read_xlsx("C:/Users/seanl/Desktop/Spring 2021/Thesis/Streptococcus sp_FDAARGOS 522 and metadata.xlsx") %>%
  janitor::clean_names() %>%
  dplyr::select(taxonomy, streptococcus_sp_fdaargos_522) %>%
  rename(sample_id = taxonomy,
         taxa = streptococcus_sp_fdaargos_522)

daargos_merge <- merge(daargos, new_metadata, by="sample_id") %>%
  mutate(taxa_bin = case_when(
    taxa > 0 ~ "1",
    taxa == 0 ~ "0"
  ))

daargos_merge$bmi <- scale(daargos_merge$bmi)
daargos_merge$age_at_a1 <- scale(daargos_merge$age_at_a1)

daargos_merge$taxa_bin <- as.numeric(daargos_merge$taxa_bin)

taxa_model <- glm(taxa_bin ~ mecfs_status + ibs_dx + site + sex + age_at_a1 + bmi + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + race_hispanic + mecfs_status*ibs_dx, family = "binomial", data = daargos_merge)

summary(taxa_model)

table(daargos_merge$taxa_bin)

daargos_cluster1 <- daargos_merge %>%
  filter(pam2 == "1")

daargos_cluster2 <- daargos_merge %>%
  filter(pam2 == "2")

taxa_model_clust1 <- glm(taxa_bin ~ mecfs_status + ibs_dx + site + sex + age_at_a1 + bmi + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + race_hispanic, family = "binomial", data = daargos_cluster1)

summary(taxa_model_clust1)

taxa_model_clust2 <- glm(taxa_bin ~ mecfs_status + ibs_dx + site + sex + age_at_a1 + bmi + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + race_hispanic, family = "binomial", data = daargos_cluster2)

summary(taxa_model_clust2)

tbl1 <- tableby(taxa_bin ~ mecfs_status + ibs_dx + site + pam2 + sex + age_at_a1 + bmi + antibiotic_use + probiotic_supp_use + prebiotic_supp_use + race_hispanic, data = daargos_merge)

summary(tbl1)

tbl1 <- as.data.frame(tbl1)
tab_df(tbl1)
tbl1

```

Demographics by FDAARGOS 522 presence
```{r fdaargos table}
daargos_merge$taxa_bin <- as.factor(daargos_merge$taxa_bin)

label(daargos_merge$sample_type) <- "ME/CFS Status"
label(daargos_merge$sex) <- "Sex"
label(daargos_merge$bmi_cat) <- "BMI Category"
label(daargos_merge$race_hispanic) <- "Race"
label(daargos_merge$ibs_dx) <- "IBS Status"
label(daargos_merge$pain_narc_rx) <- "Pain Narcotic Rx"
label(daargos_merge$antibiotic_use) <- "Antibiotic Use"
label(daargos_merge$probiotic_supp_use) <- "Probiotic Supplements"
label(daargos_merge$prebiotic_supp_use) <- "Prebiotic Supplements"
label(daargos_merge$site) <- "Site"
label(daargos_merge$pam2) <- "Cluster"
label(daargos_merge$taxa_bin) <- "Streptococcus FDAARGOS522 Presence"

getTable1Stats <- function(x, digits = 0, ...){
  getDescriptionStatsBy(x = x,
                        by = daargos_merge$taxa_bin,
                        digits = digits,
                        header_count = TRUE,
                        statistics = TRUE,
                        ...)
}

mergeDesc(getTable1Stats(daargos_merge$sex),
          getTable1Stats(daargos_merge$bmi_cat),
          getTable1Stats(daargos_merge$race_hispanic),
          getTable1Stats(daargos_merge$ibs_dx),
          getTable1Stats(daargos_merge$site),
          getTable1Stats(daargos_merge$pain_narc_rx),
          getTable1Stats(daargos_merge$pam2),
          getTable1Stats(daargos_merge$antibiotic_use),
          getTable1Stats(daargos_merge$probiotic_supp_use),
          getTable1Stats(daargos_merge$prebiotic_supp_use),
          getTable1Stats(daargos_merge$sample_type),
          htmlTable_args = list(caption = "FDAARGOS522 Presence"))

```
